### 2012-09-28 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
working on the numerical experiments section

\subsection{Parallel Adjoint Method}
\label{subsec:parallel_adjoint}
We can take much of what was learned from the survey of parallel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers. Direct analogs
can be derived from these works by noting that the primary difference
between solving a linear system with Monte Carlo methods and fixed
source Monte Carlo transport problems is the content of Markov chains
that are generated. These transitions represeneted by these chains are
bound by probabilities and weights and are initiated by the sampling
of a source. In the context of transport problems, those transitions
represent events such as pariticle scattering and absorbition with
probabilities that are determined by physical data in the form of
cross sections. For stochastic matrix inversion, those transitions
represent moving between the equations of the linear system (and
therefore the physical domain which they represent) and their
probabilites are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains.

Therefore, parallel methods for Monte Carlo radiation transport and
abstract those that apply to matrix inversion methods as an initial
means of developing a parallel Neumann-Ulam-type solver. Based on the
results observed in the last decade of parallel Monte Carlo
development, we can generate many practical questions that this type
of work could answer. Given a decomposed domain, it is clear that
domain overlap significantly reduces the amount of communication
required between adjacent domains as histories move to states that are
not owned by the local processor. How much domain overlap is suitable
for matrix inversion problems? Are we memory limited by large problems
such that only so much overlap is feasible? Are there metrics related
to the properties of the matrix such that we can provide guidelines
for selecting the amount of overlap required? Furthermore, as Wagner
and colleagues observed \citep{wagner_hybrid_2010}, with marginal
domain overlap for reactor problems, the percentage of histories
leaving the local domain can easily be reduced to less than 1\%. Evans
and colleagues in their initial MCSA development typically used only
50 particle histories in order to compute a correction in an MCSA
iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
that correlates to a statistical uncertainty of 14.1\% in the
correction yet as compared to convential Krylov solvers, MCSA achieved
the identical numerical results. If good convergence and numerically
accurate solutions can be achieved with such a large uncertainty in
the correction computation, then perhaps with enough domain overlap
the minimal amount of histories that do transition to non-local states
can be ignored and thus eliminate all communication in the parallel
Monte Carlo adjoint solver and create an embarassingly parallel
method. It will be important to determine if the expectation value
bias in the MCSA correction generated by this approximation will
remain within the bounds of the already high statistical uncertainty
for unbiased estimates, thus providing numerically equivalent results.

>>>>

\subsection{Parallel Adjoint Method}
\label{subsec:parallel_adjoint}
We can take much of what was learned from the survey of parallel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers. Direct analogs
can be derived from these works by noting that the primary difference
between solving a linear system with Monte Carlo methods and fixed
source Monte Carlo transport problems is the content of the Markov
chains that are generated. These transitions represeneted by these
chains are bound by probabilities and weights and are initiated by the
sampling of a source. In the context of transport problems, those
transitions represent events such as pariticle scattering and
absorbition with probabilities that are determined by physical data in
the form of cross sections. For stochastic matrix inversion, those
transitions represent moving between the equations of the linear
system (and therefore the physical domain which they represent) and
their probabilites are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. Based on the results observed in
the last decade of parallel Monte Carlo development, we can generate
many practical questions that this type of work could answer.

Given a decomposed domain, it is clear that domain overlap
significantly reduces the amount of communication required between
adjacent domains as histories move to states that are not owned by the
local processor. How much domain overlap is suitable for matrix
inversion problems? Are we memory limited by large problems such that
only so much overlap is feasible? Are there metrics related to the
properties of the matrix such that we can provide guidelines for
selecting the amount of overlap required? Furthermore, as Wagner and
colleagues observed \citep{wagner_hybrid_2010}, with marginal domain
overlap for reactor problems, the percentage of histories leaving the
local domain can easily be reduced to less than 1\%. Evans and
colleagues in their initial MCSA development typically used only 50
particle histories in order to compute a correction in an MCSA
iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
that correlates to a statistical uncertainty of 14.1\% in the
correction, yet as compared to convential Krylov solvers, MCSA
achieved identical numerical results. If good convergence and
numerically accurate solutions can be achieved with such a large
uncertainty in the correction computation, then perhaps with enough
domain overlap the minimal amount of histories that do transition to
non-local states can be ignored and thus eliminate all communication
in the parallel Monte Carlo adjoint solver and create an embarassingly
parallel method. It will be important to determine if the expectation
value bias in the MCSA correction generated by this approximation will
remain within the bounds of the already high statistical uncertainty
for unbiased estimates, thus providing numerically equivalent results.

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
working on the numerical experiments section

In our experimental framework, we will formulate all of our problems
using the weak form of the differential equations. Doing so allows us
to leverage modern finite element assembly engine software with
embedded automatic differentiation technology. Using the Panzer
library in the Trilinos sceintific computing framework
\citep{heroux_overview_2005}, function evaluation routines in the form
of Eq~(\ref{eq:fe_weak_form}) are generated to represent the finite
element problem and boundary conditions. Per Bartlett's automatic
differentiation work and an associated parallel linear algebra
framework, the equation sets for all physics implemented are assembled
into the nonlinear system. By leveraging the general linear algebra
framework, we are able to abstract away those concepts in the
stochastic solvers implementation and instead focus on an
algorithm-oriented approach for the solver implementation
\citep{musser_algorithm-oriented_1994}.

@@

\subsection{FANM Verification}
\label{subsec:fanm_verification}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a recitilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows :
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
pressure of the fluid and $T$ the temperature of the
fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport. In addition, we
close the system with the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the fluid pressure, $\mu$
is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux in
the fluid, and $k$ is the thermal conductivity of the fluid. We will
then apply these equations to the following three benchmark problems.

\subsubsection{Thermal Convection Cavity Problem}
\label{subsubsec:natural_convection_cavity}

\subsubsection{Lid Driven Cavity Problem}
\label{subsubsec:lid_driven_cavity}

\subsubsection{Backward-Facing Step Problem}
\label{subsubsec:backward_facing_step}

\section{Numerical Experiments}
\label{sec:numerical_experiments}
Using the verification problems for the linear and nonlinear solvers
as a testing environment, we propose a set of experiments that will be
used to characterize several qualities of these stochastic methods.

>>>>

In our experimental framework, we will formulate all of our problems
using the weak form of the differential equations. Doing so allows us
to leverage modern finite element assembly engine software with
embedded automatic differentiation technology. Using the Panzer
library in the Trilinos sceintific computing framework
\citep{notz_graph-based_2010,heroux_overview_2005}, function
evaluation routines in the form of Eq~(\ref{eq:fe_weak_form}) are
generated to represent the finite element problem and boundary
conditions. Per Bartlett's automatic differentiation work and an
associated parallel linear algebra framework, the equation sets for
all physics implemented are assembled into the nonlinear system. By
leveraging the general linear algebra framework, we are able to
abstract away those concepts in the stochastic solvers implementation
and instead focus on an algorithm-oriented approach for the solver
implementation \citep{musser_algorithm-oriented_1994}.

@@

\subsection{FANM Verification}
\label{subsec:fanm_verification}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a recitilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows :
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
pressure of the fluid and $T$ the temperature of the
fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters withing these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems.

\subsubsection{Thermal Convection Cavity Problem}
\label{subsubsec:natural_convection_cavity}
In 1983 benchmark solution for the natural convection of air in a
square cavity was published \citep{de_vahl_davis_natural_1983} as
shown in Figure~\ref{fig:natural_convection_cavity} for the solution
of the energy, mass, and momentum equations.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/research_proposal/natural_convection_cavity.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the natural convection cavity
      benchmark.} \textit{Dirichlet conditions are set for the
      temperature on the left and right while Neumann conditions are
      set on the top and bottom of the Cartesian grid. The temperature
      gradients will cause bouyancy-driven flow. Zero velocity
      Dirichlet conditions are set on each boundary. No thermal source
      was present.}}
  \label{fig:natural_convection_cavity}
\end{figure}
In this problem, a rectilinear grid is applied to the unit square. No
heat flow is allowed out of the top and bottom of the square with a
zero Neumann condition specified. Bouyancy driven flow is generated by
the temperature gradient from the cold and hot Dirichlet conditions on
the left and right boundaries of the box. By adjusting the Rayleigh
number of the fluid (and therefore adjusting the ratio of convective
to conductive heat transfer), we can adjust the influence of the
nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
on a $100 \times 100$ square mesh grid.

\subsubsection{Lid Driven Cavity Problem}
\label{subsubsec:lid_driven_cavity}
As an extension of the convection problem, the second benchmark
problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
flow to introduce higher Reynolds numbers into the system, providing
more intertial force to overcome the viscous forces in the fluid. The
setup for this problem is equally simple, containing only the
Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
and is only applied to the mass and momentum equations on the unit
square.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/research_proposal/lid_driven_cavity.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the lid driven cavity benchmark.}
    \textit{Dirichlet conditions of zero are set for the velocity on
      the left and right and bottom while the Dirichlet condition set
      on the top provides a driving force on the fluid.}}
  \label{fig:lid_driven_cavity}
\end{figure}
The top boundary condition will provide a driver for the flow and its
variation will in turn vary the Reynolds number of the fluid. An
increased velocity will generate more inertial forces in the fluid,
which will overcome the viscous forces and again increase the
influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
also used a $100 \times 100$ grid for this benchmark problem with
Reynolds numbers up to \sn{1}{4}.

\subsubsection{Backward-Facing Step Problem}
\label{subsubsec:backward_facing_step}
The third benchmark was generated by Gartling in 1990 and consists of
both flow over a backward step and an outflow boundary condition
\citep{gartling_test_1990}. Using the mass and momentum equations
while neglecting the energy equation, this problem utilizes a longer
domain with a 1/30 aspect ratio with the boundary conditions as shown
in Figure~\ref{fig:backward_facing_step}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.3}{
      \input{chapters/research_proposal/backward_facing_step.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for the backward facing step
      benchmark.} \textit{Zero velocity boundary conditions are
      applied at the top and bottom of the domain while the outflow
      boundary condition on the right boundary is represented by zero
      stress tensor components in the direction of the flow. For the
      inlet conditions, the left boundary is split such that the top
      half has a fully formed parabolic flow profile and the bottom
      half has a zero velocity condition, simulating flow over a
      step.}}
  \label{fig:backward_facing_step}
\end{figure}
In this problem, the inflow condition is specified by a fully-formed
parabolic flow profile over a zero velocity boundary representing a
step. The flow over this step will generate a recirculating backflow
under the inlet flow towards the step. As in the lid driven cavity
problem, the nonlinear behavior of this benchmark and the difficulty
in obtaining a solution is dictated by the Reynolds number of the
fluid. In Shadid's work, a $20 \times 400$ non-square recitlinear grid
was used to discretize the domain with Reynolds number up to
\sn{5}{2}.

\section{Proposed Numerical Experiments}
\label{sec:numerical_experiments}
Using the verification problems for the linear and nonlinear solvers
as a testing environment, we propose a set of experiments that will be
used to characterize several qualities of these stochastic methods. At
the end of Chapter~\ref{ch:stochastic_methods}, we posed several
research questions related to the parallelization of the adjoint
Neumann-Ulam method and the MCSA method while in
Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
the FANM method. We will generate numerical experiments from the these
questions by dividing them into 4 categories: domain overlap studies
for the parallel Neumann-Ulam method, domain replication studies for
the parallel Neumann-Ulam method, parallel performance and numerical
accuracy studies for the MCSA method, and parallel performance and
numerical accuracy studies for the FANM method.

\subsection{Domain Overlap Studies for the Parallel Neumann-Ulam
  Method}
\label{subsec:domain_overlap_studies}
Characterizing and understanding the domain overlap behavior of a
parallel Neumann-Ulam method will be critical to efficient
implementations of other solvers that will be built on top of these
methods. In the reactor physics work that we reviewed, the amount
domain overlap used could be directly correlated to physical
parameters of the system, namely the dimension of repeated geometric
structures and the mean free path of neutrons in the system. We
therefore seek analgous measurements for the general linear
system. These measurements will characterize how far from its starting
state a random walk sequence moves as a function of properties of the
linear system. Our experimental framework will allow us to fix various
properties of the system while varying others including the size of
the linear system, its spectral radius, and other parameters related
to the Eigenvalues and conditioning of the linear operator. By
varying the domain overlap and measuring parallel performance with
respect to these properties, we can develop guidelines for setting the
overlap with respect to measurable properties of the linear
operator. In addition, varying the amount of overlap will allow us to
measure the additional memory costs incurred.

\subsection{Domain Replication Studies for the Parallel Neumann-Ulam
  Method}
\label{subsec:domain_replication_studies}
Employing domain replication with the Neumann-Ulam method will be an
exercise in batch statistics. By using replication, we also expect to
incur an additional memory overhead. As noted in previous work, domain
replication was coupled with domain overlap in order to improve load
balancing and scalability in the implementations. As replication will
likely be required to solve future resiliency issues and may even by
required by this work in order to achieve scalability, it parallel
performance must also be measured. We will therefore measure the
memory and performance overhead incurred by varying the amount of
domain replication leveraged by the Neumann-Ulam solver with respect
to a fixed linear system.

\subsection{Parallel Performance and Numerical Accuracy Studies for MCSA
  Method}
\label{subsec:mcsa_studies}
A parallel MCSA implementation and our numerical test framework will
allow us to compare its performance to Krylov methods and allow us to
experiment with the way we combine it with the Neumann-Ulam solver. As
disucussed in \S~\ref{subsec:parallel_adjoint}, previous work with
MCSA used a correction generated by the Neumann-Ulam method with a
fairly large uncertainty near 14\%. We hypothesized that if we can
identify the amount of overlap in or Neumann-Ulam experiments required
to eliminate most transitions between adjacent domains owned by
different processors, then perhaps transport to adjacent domains can
be ignored. This will result in a biased estimator for the MCSA
correction that cannont be reproduced, however, these studies will aim
to determine if that bias and lack of reproducablity affects the
numerical results and reproducability of the MCSA method. Furthermore,
we will use the experimental framework to measure the convergence
rates for nonsymmetric systems and memory usage of the MCSA algorithm
and compare them to conventional Krylov methods using the same
framework.

\subsection{Parallel Performance and Numerical Accuracy Studies for FANM
  Method}
\label{subsec:fanm_studies}
